A Monte Carlo Synthetic-Acceleration Method for Solving the Thermal Radiation Diffusion Equation

### 2012-07-19 doc/jcomp2012/jcomp2012.tex
first draft done for mcsa jcomp

%%---------------------------------------------------------------------------%%
%% Monte Carlo Synthetic Acceleration Paper
%% Submitted to Journal of Computational Physics 2012
%%---------------------------------------------------------------------------%%
%%
%% Copyright 2007, 2008, 2009 Elsevier Ltd
%%
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%%

@@

  \title{A Monte Carlo Synthetic Acceleration Method for Solving the
    Thermal Radiation Diffusion Equation}

@@

  \address[ornl]{Oak Ridge National Laboratory, 
    1 Bethel Valley Rd., Oak Ridge, TN 378131 U.S.A.}
  \address[wisc]{University of Wisconsin-Madison, 1500 Engineering
    Dr., Madison, WI 53716 U.S.A.}

@@

    We present a synthetic-acceleration based Monte Carlo method for
    solving the 1T thermal radiation diffusion equation.  We show that
    this method can be an effective solver for sparse matrix systems
    and is competitive with deterministic methods.  We also discuss
    its general applicability to broader classes of problems.

@@

The purpose of this study is to demonstrate an efficient Monte Carlo
solution method for 3D, time-dependent discrete systems. In this
work, we extend our previous work in the following manner:

\begin{itemize}
\item We develop a rigorous synthetic-acceleration Monte Carlo method
  for solving sparse matrix systems.
\item We apply this solver to the 1T thermal radiation diffusion
  equation in 3D.
\end{itemize}

In \S~\ref{sec:monte-carlo-matrix} we review Monte Carlo methods for
solving matrix systems. In \S~\ref{sec:iter-refin-monte} we discuss
our new synthetic-acceleration scheme and connect with previous work
by Halton~\cite{halton_1994} and others~\cite{evans_2003}.  The 1T
radiation diffusion equation and its discrete form will be derived in
\S~\ref{sec:discr-form-radi}.  We discuss using Synthetic-Accleration
Monte Carlo to solve the discrete radiation diffusion equation in
\S~\ref{sec:solut-radi-diff}.  Results from two 3D test problems are
given in \S~\ref{sec:results}.  We finish with conclusions and ideas
for future work in \S~\ref{sec:conclusions}.

@@

The terms in the Neumann series can be interpreted as a series of
transitions from $i_{m-1}\rightarrow i_m$ that can be simulated by a
random walk.  Let $X$ be a random variable sampled from a random walk
with $k$ events (particles) that initiates in state $i$,

@@

Here, the particle weight on the $m^\text{th}$ step is denoted $W_m$,
and each particle starts with unit weight. At every step,
contributions to each component of $X$ are provided only by the
starting state of the particle, $b_{i_m}$. Therefore, for a given
random walk permutation, each state must be used as a starting value
in order to calculated its contribution to the expectation value
(i.e. a state of size $N$ will have $N$ particles per random walk
permutation). When the particle transitions between states, for
example $i\rightarrow j$, the particle weight is multiplied by the
transition factor.

@@

We are left to define the transition probabilities, $p_{ij}$. The most
straightforward approach is to set

@@

Each row of the transition probability matrix, $\vP$, represents a
discrete probability density function that can be sampled to select a
new state $j$, given that the current state is $i$.  Random walks can
be terminated in two ways: the matrix can be augmented with a
terminating event equation that describes the probability that a
particle ends its walk, or the random walk can be terminated by weight
cutoff, $W_c$.  Generally, we choose to terminate random walks using a
weight cutoff as opposed to augmenting the matrix sucth that the
random walk terminates when $W_m < W_c$.

@@

Similar to the Direct Method, the Adjoint Method random walk process
requires a terminating condition.  In all of the work that follows we
utilize a relative weight cutoff.  The weight cutoff is defined as a
fraction of the starting weight such that

@@

where $W_f$ is the terminating weight of the random walk, $W_c$ is the
input relative weight cutoff, and $\hat{b}_{i_0}$ is the initial
weight of the source particle.  The random walk terminates on the
$m^\text{th}$ step if $W_m < W_f$.

%%---------------------------------------------------------------------------%%
\section{Synthetic-Acceleration Monte Carlo}
\label{sec:iter-refin-monte}

@@

We propose a modification of this scheme that uses Monte Carlo as a
synthetic-acceleration scheme for fixed-point iteration. We begin by
subtracting Eq.~(\ref{eq:richardson}) from Eq:~(\ref{eq:point_iteration}):

@@

is defined as the error at iteration $l+1$. We then subtract
$(\vI-\vA) \delta \vx^{l+1}$ from Eq.~(\ref{eq:synthetic_error}) giving:

\begin{equation}
  \vA \delta \vx^{l+1} = (\vI - \vA)(\vx^{l+1} - \vx^{l})\:.
  \label{eq:synthetic_error2}
\end{equation}

The following iterative scheme will then converge in one iteration:

\begin{subequations}
  \begin{gather}
    \vx^{l+1} = (\vI - \vA)\vx^l + \vb\:,\\
    %%
    \vA \delta \vx^{l+1} = (\vI - \vA)(\vx^{l+1} - \vx^{l})\:.
    \label{eq:synthetic-residual_solve}\\ 
    %% 
    \vx=\vx^{l+1}+\delta\vx^{l+1}\:.
  \end{gather}
  \label{eq:general-synthetic}
\end{subequations}

To accelerate the fixed point method, we can instead create a new
iterative scheme by approximating the operator $\vA$ using the Monte
Carlo method to estimate the solution error. With this idea hand, the
Fixed-Point Monte Carlo Synthetic-Acceleration (MCSA) method can be
defined as follows:

@@

The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that
the Monte Carlo solution only approximately inverts this operator.
However, $\hat{\vA}$ is the same operator as $\vA$;
$\hat{\vA}\equiv\vA$.  Thus, we have defined a scheme in which the
initial estimate of $\vx$ in each iteration is updated using a single
fixed-point iteration\footnote{Although, a Krylov method could be used
  here.}.  The residual is calculated and is used as a source to
estimate the error, $\delta\vx^{l+1/2}$, by solving
Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint
method. The error is then used to calculate an updated iterate of the
solution vector, $\vx^{l+1}$.  The entire sequence is iterated to
convergence based on the following stopping
criterion~\cite{kelley_1995},
\begin{equation}
  \|\vr\|_{\infty} < .pdfilon\cdot\|\vb\|_\infty\:.
  \label{eq:stopping-criteria}
\end{equation}

@@

A timing study and convergence study is used to demonstrate the
effectiveness of the adjoint method as compared to the direct
method. To assess both the CPU time and number of iterations required
to converge to a solution, a problem of constant $\Delta$ and was used
with varying values of grid size, fixing the spectral radius of the
system at a constant value for each variation. Both the five point and
nine point stencils were used with both the direct and adjoint
solvers. For each case, 50 random walk permuations were simulated per
iteration with a weight cutoff of $1.0E-4$ for the stochastic linear
solver and a convergence tolerance of $1.0E-8$ for the MCSA iterative
solver. All computations were completed on a 2.66 GHz Intel Core i7
machine with 8 GB 1067 MHz DDR3 memory used for all computations
presented in this section. Fig.~\ref{fig:poisson_cpu_time} gives the
CPU time needed for each case to converge in seconds and
Fig.~\ref{fig:poisson_iterations} gives the number of iterations
needed for each case to converge relative to the problem size.

\begin{figure}[htpb!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time.pdf}
  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
    $N \times N$ mesh). Both the adjoint and direct solvers are used
    with the five point and nine point stencils. A significant CPU
    time speedup is noted with the adjoint method due to the reduced
    number of histories.}
  \label{fig:poisson_cpu_time}
\end{figure}

\begin{figure}[htbp!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_Iterations.pdf}
  \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
    \times N$ mesh. Both the adjoint and direct solvers are used with
    the five point and nine point stencils. Both methods give similar
    iteration behavior.}
  \label{fig:poisson_iterations}
\end{figure}

We see clearly in Fig.~\ref{fig:poisson_cpu_time} that the using the
adjoint solver with MCSA results in several orders of magnitude
speedup over the direct solver while the number of iterations required
to converge is of a similar scale. We expect this for several
reasons. First, with an equivalent number of histories specified for
both solvers and a system of size $N \times N$, the direct solver will
compute $N \times N$ random walks per permuation while the adjoint solver
will only compute one. Initially, this may not seem like a fair
comparison, however, we see from Fig.~\ref{fig:poisson_cpu_time} that
the number of iterations requried to converge is approximately the
same. This is necessary in the direct method to ensure a contribution
from each state as the random walk sequence will only contribute to
the starting state. For the adjoint method, because the random walk
sequence contributes to the state in which it currently resides, fewer
histories are neccessary to compute contributions from all the nonzero
parts of the source. Therefore, although the total number of random
walks per iteration is in fact larger for the direct method, the
solution computed at each iteration is no better than that computed by
the adjoint method.

@@

\begin{figure}[htpb!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_Convergence.pdf}
  \caption{\sl Infinity norm of the solution residual vs. iteration
    number for a problem of fixed size. Both the adjoint and direct
    solvers are used with the five point and nine point stencils. Both
    methods give similar iteration behavior.}
  \label{fig:poisson_convergence}
\end{figure}

@@

\begin{figure}[htpb!]
  \centerline{
    \input{mesh_cell.pdftex_t}}
  \caption{Orthogonal 3D mesh Cell.}
  \label{fig:cell}
\end{figure}

@@

In the next section we will derive an MCSA method to solve this
equation.

@@

\begin{figure}[htpb!]
  \centerline{
    \input{3D_mesh_segment.pdftex_t}}
  \caption{3D mesh-segment.  The particle's current cell (in blue) is
    connected to 6 adjacent cells.  The cell numbers are plotted
    alongside the mesh such that the blue cell has index 7, the high
    $z$ cell has index 2, etc.}
  \label{fig:mesh_segment}
\end{figure}

@@

\begin{figure}[htpb!]
  \centerline{
    \includegraphics[width=5in,clip]{marshak_10.pdf}}
  \caption{
    Solution of the Marshak wave problem at $t=10$~sh.  The analytic
    solution is compared to the MCSA solution.  Note that the MCSA,
    PCG, and PFIXED methods all give numerically identical answers.}
  \label{fig:marshak_10}
\end{figure}

@@

\begin{figure}[htpb!]
  \centerline{
    \includegraphics[width=5in,clip]{spectral_radius.pdf}}
  \caption{Spectral radius of the iteration matrix, $\ve{H}$, for the
    Marshak Wave problem as a function of elapsed problem time.}
  \label{fig:marshak_spectral}
\end{figure}

@@

\begin{figure}[htpb!]
  \centerline{
    \includegraphics[width=5in,clip]{mrshk_np_CPU.pdf}}
  \caption{CPU time versus number of particles per stage for the
    Marshak problem.  The number of particles is the requested number.
    In each stage the exact number can vary depending upon the number
    of cells and source strength per cell.}
  \label{fig:CPU_Np}
\end{figure}

\begin{figure}[htpb!]
  \centerline{
    \includegraphics[width=5in,clip]{mrshk_np_Ns.pdf}}
  \caption{
    Maximum number of stages per cycle versus number of particles per
    stage for the Marshak problem.  The number of particles is the
    requested number.  In each stage the exact number can vary
    depending upon the number of cells and source strength per cell.}
  \label{fig:CPU_Ns}
\end{figure}

@@

\begin{figure}[htpb!]
  \centerline{
    \includegraphics[width=5in,clip]{mrshk_wc_CPU.pdf}}
  \caption{
    CPU time versus relative weight cutoff for the Marshak problem.
    Results are shown for 10 and 50 requested particles per stage.
    The numbers in parenthesis show the maximum number of particles
    run in a stage during the course of the problem.}
  \label{fig:CPU_wc}
\end{figure}

@@

\begin{table}[htpb!]
  \begin{center}
    \begin{tabular}{lrr}\hline\hline
      \multicolumn{1}{c}{Method} & 
      \multicolumn{1}{c}{Max Iterations} & 
      \multicolumn{1}{c}{Relative CPU Time}\\\hline\hline
      %%
      PCG & 11 & 1.11 \\
      %%
      PFIX & 11 & 1.42 \\
      %%
      MCSA & 11 & 1.0 \\
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{Comparison of solution methods for the Marshak problem. All
    methods were run with a stopping criterion of $\sn{1}{-8}$.  The
    MCSA method used $N_p=10$ and $W_c = \sn{1}{-4}$.  The CPU time is
    measured relative to the MCSA method, which required 72.6 s on a
    3.6~GHz Xeon.}
  \label{tab:marshak_comparison}
\end{table}

@@

\begin{figure}[htpb!]
  \centerline{
    \includegraphics[width=5in,clip]{mesh.pdf}}
  \caption{Multi-material problem mesh.  The duct region is blue.  The
    wall is red.  The foil is green.  The mesh is $60\times 60\times
    60$.}
  \label{fig:multi-mat-mesh}
\end{figure}

\begin{figure}[htpb!]
  \centerline{
    \includegraphics[width=5in,clip]{T_multi_mat.pdf}}
  \caption{
    Two-dimensional plot of the temperature at 100~sh on a cut-plane
    positioned at the mid-point of the $z$-axis.  The positions of the
    edit points where the time-evolution of the temperature is plotted
    in Fig.~\ref{fig:edit-points} are indicated by the black disks.}
  \label{fig:multi-mat_T}
\end{figure}

\begin{figure}[htpb!]
  \begin{center}
    \begin{tabular}{cc}
      \subfigure[Edit points]{\includegraphics[width=1.5in,clip]{points.pdf}} &
      \subfigure[Temperature]{\includegraphics[width=3.5in,clip]{multi_mat.pdf}}
    \end{tabular}
  \end{center}
  \caption{(a) Edit points and (b) time-evolution of the temperature
    at each point.  All points are centered in the $z$-axis.}
  \label{fig:edit-points}
\end{figure}

@@

\begin{table}[htpb!]
  \begin{center}
    \begin{tabular}{lrr}\hline\hline
      \multicolumn{1}{c}{Method} & 
      \multicolumn{1}{c}{Max Iterations} & 
      \multicolumn{1}{c}{Relative CPU Time}\\\hline\hline
      %%
      PCG & 18 & 1.03 \\
      %%
      PFIX & 38 & 1.43 \\
      %%
      MCSA & 20 & 1.0 \\
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{
    Comparison of solution methods for the multi-material problem. All
    methods were run with a stopping criterion of $\sn{1}{-8}$.  The
    MCSA method used $N_p=1000$ and $W_c = \sn{1}{-3}$.  The problem
    was run to an elapsed time of 10~sh.  The MCSA method required 3.4
    hours on a 3.6~GHz Xeon.}
  \label{tab:multimat_comparison}
\end{table}

>>>>

%%---------------------------------------------------------------------------%%
%% Monte Carlo Synthetic-Acceleration Paper
%% Submitted to Journal of Computational Physics 2012
%%---------------------------------------------------------------------------%%
%%
%% Copyright 2007, 2008, 2009 Elsevier Ltd
%%
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%%

@@

  \title{A Monte Carlo Synthetic-Acceleration Method for Solving the
    Thermal Radiation Diffusion Equation}

@@

  \address[ornl]{Oak Ridge National Laboratory, 
    1 Bethel Valley Rd., Oak Ridge, TN 37831 U.S.A.}
  \address[wisc]{University of Wisconsin-Madison, 1500 Engineering
    Dr., Madison, WI 53716 U.S.A.}

@@

    We present a synthetic-acceleration based Monte Carlo method for
    solving the 1T thermal radiation diffusion equation in 3
    dimensions. We show that this method can be an effective solver
    for sparse matrix systems and is competitive with deterministic
    methods including preconditioned Conjugate Gradient. We also
    discuss various aspects of preconditioning and its general
    applicability to broader classes of problems.

@@

The purpose of this study is to demonstrate an efficient Monte Carlo
solution method for 3D, time-dependent, discrete systems. In this
work, we extend our previous work in the following manner:

\begin{itemize}
\item We develop a rigorous synthetic-acceleration Monte Carlo method
  for solving sparse matrix systems.
\item We apply this method to the 1T thermal radiation diffusion
  equation in 3D.
\end{itemize}

In \S~\ref{sec:monte-carlo-matrix} we review Monte Carlo methods for
solving matrix systems. In \S~\ref{sec:iter-refin-monte} we discuss
our new synthetic-acceleration scheme and connect with previous work
by Halton~\cite{halton_1994} and others~\cite{evans_2003}.  The 1T
radiation diffusion equation and its discrete form will be derived in
\S~\ref{sec:discr-form-radi}.  We discuss using Monte Carlo
Synthetic-Accleration to solve the discrete radiation diffusion
equation in \S~\ref{sec:solut-radi-diff}.  Results from two 3D test
problems are given in \S~\ref{sec:results}.  We finish with
conclusions and ideas for future work in \S~\ref{sec:conclusions}.

@@

The terms in the Neumann series can be interpreted as a series of
transitions from $i_{m-1}\rightarrow i_m$ that can be simulated by a
random walk.  Let $X$ be a random variable sampled from a random walk
with $k$ events (particles in the context of radiative transfer) that
initiates in state $i$,

@@

Here, the particle weight on the $m^\text{th}$ step is denoted $W_m$,
and each particle starts with unit weight. At every step,
contributions to each component of $X$ are generated only in the
starting state of the particle, $i_0$, by the source term in the
current state, $b_{i_m}$. Therefore, for a given random walk
permutation, each state must be used as a starting value in order to
calculate its component of the expectation value (i.e. a state of size
$N$ will have $N$ particles per random walk permutation). When the
particle transitions between states, for example $i\rightarrow j$, the
particle weight is multiplied by the transition factor.

@@

We are left to define the transition probabilities. The most
straightforward approach is to set

@@

Each row of the transition probability matrix, $\vP$, represents a
discrete probability density function that can be sampled to select a
new state $j$, given that the current state is $i$.  Random walks can
be terminated in two ways: the matrix can be augmented with a
terminating event equation that describes the probability that a
particle ends its walk, or the random walk can be terminated by weight
cutoff, $W_c$.  Generally, we choose to terminate random walks using a
weight cutoff as opposed to augmenting the matrix such that the random
walk terminates when $W_m < W_c$.

@@

Similar to the direct method, the adjoint method random walk process
requires a terminating condition.  In all of the work that follows we
utilize a relative weight cutoff.  The weight cutoff is defined as a
fraction of the starting weight such that

@@

where $W_f$ is the terminating weight of the random walk and $W_c$ is
the input relative weight cutoff. The random walk terminates on the
$m^\text{th}$ step if $W_m < W_f$.

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Synthetic-Acceleration}
\label{sec:iter-refin-monte}

@@

We propose a modification of this scheme that uses Monte Carlo as a
synthetic-acceleration for the fixed-point iteration given by
Eq.~(\ref{eq:richardson}). We begin by subtracting
Eq.~(\ref{eq:richardson}) from Eq:~(\ref{eq:point_iteration}):

@@

is defined as the error at iteration $l+1$. We then subtract
$(\vI-\vA) \delta \vx^{l+1}$ from Eq.~(\ref{eq:synthetic_error}),
giving a linear system to solve for the error:

\begin{equation}
  \begin{split}
    \vA \delta \vx^{l+1} &= (\vI - \vA)(\vx^{l+1} - \vx^{l}) \\
    &= \vr^{l+1}\:.
  \end{split}
  \label{eq:synthetic_error2}
\end{equation}

The following scheme will then converge in one iteration:

\begin{subequations}
  \begin{gather}
    \vx^{l+1} = (\vI - \vA)\vx^l + \vb\:,\\
    %%
    \vA \delta \vx^{l+1} = \vr^{l+1}
    \label{eq:synthetic-residual-solve}\\ 
    %% 
    \vx=\vx^{l+1}+\delta\vx^{l+1}\:.
  \end{gather}
  \label{eq:general-synthetic}
\end{subequations}

To accelerate the fixed-point method and avoid inverting $\vA$
directly in Eq.~(\ref{eq:synthetic-residual-solve}), we can instead
use the above ideas to create a new iterative scheme by using the
Monte Carlo method to estimate the solution error. With this in hand,
the Fixed-Point Monte Carlo Synthetic-Acceleration (MCSA) method can
be defined as follows:

@@

The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that
the Monte Carlo solution only approximately inverts this operator.
However, $\hat{\vA}$ is the same operator as $\vA$;
$\hat{\vA}\equiv\vA$.  Thus, we have defined a scheme in which the
initial estimate of $\vx$ in each iteration is updated using a single
fixed-point iteration\footnote{Although, a Krylov method could be used
  instead of a stationary method here.}.  The residual is calculated
and is used as a source to estimate the error, $\delta\vx^{l+1/2}$, by
solving Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint
method. The error is then used to calculate an updated iterate of the
solution vector, $\vx^{l+1}$.  The entire sequence is iterated to
convergence based on the following stopping
criterion~\cite{kelley_1995},
\begin{equation}
  \|\vr\|_{\infty} < \epsilon\cdot\|\vb\|_\infty\:.
  \label{eq:stopping-criteria}
\end{equation}

@@

A timing study and convergence study is used to demonstrate the
effectiveness of the adjoint method as compared to the direct
method. To assess both the CPU time and number of iterations required
to converge to a solution, a problem of constant $\Delta$ and was used
with varying values of grid size, fixing the spectral radius of the
system at a constant value for each variation. Both the five point and
nine point stencils were used with both the direct and adjoint
solvers. For each case, 50 random walk permuations were simulated per
iteration with a weight cutoff of $1.0E-4$ for the stochastic linear
solver and a convergence tolerance of $1.0E-8$ for the MCSA iterative
solver. All computations presented in this section were completed on a
2.66 GHz Intel Core i7 machine with 8 GB 1067 MHz DDR3 memory
. Fig.~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
case to converge in seconds and Fig.~\ref{fig:poisson_iterations}
gives the number of iterations needed for each case to converge
relative to the problem size.

\begin{figure}[h!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time.pdf}
  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
    $N \times N$ square mesh). Both the adjoint and direct solvers are
    used with the five point and nine point stencils. A significant
    CPU time speedup is noted with the adjoint method due to the
    reduced number of random walk events.}
  \label{fig:poisson_cpu_time}
\end{figure}

\begin{figure}[h!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_Iterations.pdf}
  \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
    \times N$ square mesh. Both the adjoint and direct solvers are
    used with the five point and nine point stencils. Both methods
    give similar iteration behavior.}
  \label{fig:poisson_iterations}
\end{figure}

We see clearly in Fig.~\ref{fig:poisson_cpu_time} that the using the
adjoint solver with MCSA results in several orders of magnitude
speedup over the direct solver while the number of iterations required
to converge is of a similar scale. We expect this for several
reasons. First, with an equivalent number of histories specified for
both solvers and a system of size $N \times N$, the direct solver will
compute $N \times N$ random walks per permuation while the adjoint
solver will only compute one. Initially, this may not seem like a fair
comparison, however, we see from Fig.~\ref{fig:poisson_iterations}
that the number of iterations requried to converge is approximately
the same. This is necessary in the direct method to ensure a
contribution from each state as the random walk sequence will only
contribute to the starting state. For the adjoint method, because the
random walk sequence contributes to the state in which it currently
resides, fewer histories are neccessary to compute contributions from
all the nonzero parts of the source. Therefore, although the total
number of random walks per iteration is in fact larger for the direct
method, the solution computed at each iteration is no better than that
computed by the adjoint method.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_Convergence.pdf}
  \caption{\sl Infinity norm of the solution residual vs. iteration
    number for a problem of fixed size. Both the adjoint and direct
    solvers are used with the five point and nine point stencils. Both
    methods give similar iteration behavior.}
  \label{fig:poisson_convergence}
\end{figure}

@@

\begin{figure}[h!]
  \centerline{
    \input{mesh_cell.pdftex_t}}
  \caption{Orthogonal 3D mesh cell.}
  \label{fig:cell}
\end{figure}

@@

@@

\begin{figure}[h!]
  \centerline{
    \input{3D_mesh_segment.pdftex_t}}
  \caption{3D mesh-segment.  The particle's current cell (in blue) is
    connected to 6 adjacent cells.  The cell numbers are plotted
    alongside the mesh such that the blue cell has index 7, the high
    $z$ cell has index 2, etc.}
  \label{fig:mesh_segment}
\end{figure}

@@

\begin{figure}[h!]
  \centerline{
    \includegraphics[width=5in,clip]{marshak_10.pdf}}
  \caption{
    Solution of the Marshak wave problem at $t=10$~sh.  The analytic
    solution is compared to the MCSA solution.  Note that the MCSA,
    PCG, and PFIXED methods all give numerically identical answers.}
  \label{fig:marshak_10}
\end{figure}

@@

\begin{figure}[h!]
  \centerline{
    \includegraphics[width=5in,clip]{spectral_radius.pdf}}
  \caption{Spectral radius of the iteration matrix, $\ve{H}$, for the
    Marshak Wave problem as a function of elapsed problem time.}
  \label{fig:marshak_spectral}
\end{figure}

@@

\begin{figure}[h!]
  \centerline{
    \includegraphics[width=5in,clip]{mrshk_np_CPU.pdf}}
  \caption{CPU time versus number of particles per stage for the
    Marshak problem.  The number of particles is the requested number.
    In each stage the exact number can vary depending upon the number
    of cells and source strength per cell.}
  \label{fig:CPU_Np}
\end{figure}

\begin{figure}[h!]
  \centerline{
    \includegraphics[width=5in,clip]{mrshk_np_Ns.pdf}}
  \caption{
    Maximum number of stages per cycle versus number of particles per
    stage for the Marshak problem.  The number of particles is the
    requested number.  In each stage the exact number can vary
    depending upon the number of cells and source strength per cell.}
  \label{fig:CPU_Ns}
\end{figure}

@@

\begin{figure}[h!]
  \centerline{
    \includegraphics[width=5in,clip]{mrshk_wc_CPU.pdf}}
  \caption{
    CPU time versus relative weight cutoff for the Marshak problem.
    Results are shown for 10 and 50 requested particles per stage.
    The numbers in parenthesis show the maximum number of particles
    run in a stage during the course of the problem.}
  \label{fig:CPU_wc}
\end{figure}

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{lrr}\hline\hline
      \multicolumn{1}{c}{Method} & 
      \multicolumn{1}{c}{Max Iterations} & 
      \multicolumn{1}{c}{Relative CPU Time}\\\hline\hline
      %%
      PCG & 11 & 1.11 \\
      %%
      PFIX & 11 & 1.42 \\
      %%
      MCSA & 11 & 1.0 \\
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{Comparison of solution methods for the Marshak problem. All
    methods were run with a stopping criterion of $\sn{1}{-8}$.  The
    MCSA method used $N_p=10$ and $W_c = \sn{1}{-4}$.  The CPU time is
    measured relative to the MCSA method, which required 72.6 s on a
    3.6~GHz Xeon.}
  \label{tab:marshak_comparison}
\end{table}

@@

\begin{figure}[h!]
  \centerline{
    \includegraphics[width=5in,clip]{mesh.pdf}}
  \caption{Multi-material problem mesh.  The duct region is blue.  The
    wall is red.  The foil is green.  The mesh is $60\times 60\times
    60$.}
  \label{fig:multi-mat-mesh}
\end{figure}

\begin{figure}[h!]
  \centerline{
    \includegraphics[width=5in,clip]{T_multi_mat.pdf}}
  \caption{
    Two-dimensional plot of the temperature at 100~sh on a cut-plane
    positioned at the mid-point of the $z$-axis.  The positions of the
    edit points where the time-evolution of the temperature is plotted
    in Fig.~\ref{fig:edit-points} are indicated by the black disks.}
  \label{fig:multi-mat_T}
\end{figure}

\begin{figure}[h!]
  \begin{center}
    \begin{tabular}{cc}
      \subfigure[Edit points]{\includegraphics[width=2in,clip]{points.pdf}} &
      \subfigure[Temperature]{\includegraphics[width=4in,clip]{multi_mat.pdf}}
    \end{tabular}
  \end{center}
  \caption{(a) Edit points and (b) time-evolution of the temperature
    at each point.  All points are centered in the $z$-axis.}
  \label{fig:edit-points}
\end{figure}

@@

\begin{table}[htbp!]
  \begin{center}
    \begin{tabular}{lrr}\hline\hline
      \multicolumn{1}{c}{Method} & 
      \multicolumn{1}{c}{Max Iterations} & 
      \multicolumn{1}{c}{Relative CPU Time}\\\hline\hline
      %%
      PCG & 18 & 1.03 \\
      %%
      PFIX & 38 & 1.43 \\
      %%
      MCSA & 20 & 1.0 \\
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{
    Comparison of solution methods for the multi-material problem. All
    methods were run with a stopping criterion of $\sn{1}{-8}$.  The
    MCSA method used $N_p=1000$ and $W_c = \sn{1}{-3}$.  The problem
    was run to an elapsed time of 10~sh.  The MCSA method required 3.4
    hours on a 3.6~GHz Xeon.}
  \label{tab:multimat_comparison}
\end{table}
